---
title: "SAGs - Streamlining"
author: "Aleix Obiol"
date: "2026-09-14"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
source("../aux_functions_sags.R")
library(ggtreeExtra)
```

## Links to ref genomes

### Prymnesiophyceae

    - *Phaeocystis globosa*: https://ngdc.cncb.ac.cn/gwh/search/advanced/result?search_category=&search_term=&source=0&query_box=phaeocystis%20globosa
    - *Chrysotila carterae*: https://ftp.ncbi.nlm.nih.gov/genomes/all/GCA/054/660/715/GCA_054660715.1_Ccar_1.0/
    - *Prymnesium parvum*: https://ftp.ncbi.nlm.nih.gov/genomes/all/GCA/040/529/105/GCA_040529105.1_pryPar1.h1/ 
    - *Gephyrocapsa huxleyi*: https://ftp.ncbi.nlm.nih.gov/genomes/all/GCF/000/372/725/GCF_000372725.1_Emiliana_huxleyi_CCMP1516_main_genome_assembly_v1.0/
    - *Calcidiscus leptoporus*: https://ftp.ncbi.nlm.nih.gov/genomes/all/GCA/057/557/405/GCA_057557405.1_RBI_CaLept/

### Chlorarachnea

    - *Bigelowiella natans*: https://ftp.ncbi.nlm.nih.gov/genomes/all/GCA/000/320/545/GCA_000320545.1_Bigna1/

### Chrysophyceae

    - *Chrysophyceae sp. TOSAG23-4*: https://www.genoscope.cns.fr/tara/
    - *Chrysophyceae sp. TOSAG23-5*: https://www.genoscope.cns.fr/tara/ 
    - *Ochromonadaceae sp. CCMP2298*: https://www.ncbi.nlm.nih.gov/datasets/genome/GCA_026770595.1/

## Read data

```{r}
pf_tree <- 
  read.tree('../../data/trees/PF.SAGs.LGC60GCF-PMSF.treefile')

metadata_tree_raw <-
  read_tsv("../../data/trees/tree_metadata.tsv")

sags_metadata <- 
  readxl::read_excel("../../data/sags/Table-S2.xlsx") |>
  select(
    sag_id = 1,
    estimated_genome_size = `Estimated size (Mb)`,
    completeness = `Completeness (%) (out of 255)`,
    perc_intronic = `Percent Intronic`,
    perc_exonic = `Percent Exonic`,
    perc_intergenic = `Percent Intergenic`
  ) |> 
  left_join(metadata_tree_raw |> select(id, sag_id)) |> 
  select(id, everything())

additional_genomes_busco <- 
  list.files('../../data/sags/ref_genomes', pattern = '^busco_odb10_', full.names = T) |> 
  map(read_tsv) |>
  bind_rows() |> 
  mutate(species = str_replace_all(str_remove(Input_file, '\\.fasta'), '_', ' ')) |> 
  left_join(metadata_tree_raw |> select(id, species))

additional_genomes_stats <- 
  read_tsv(
    '../../data/sags/ref_genomes/ref_genomes_stats.tsv'
  ) |> 
  mutate(species = str_replace_all(str_remove(file, '\\.fasta'), '_', ' ')) |> 
  left_join(metadata_tree_raw |> select(id, species))  

additional_genomes_architecture <- 
  read_tsv('../../data/sags/ref_genomes/Table_genome_prediction_stats_genome_reference.tsv') |> 
  mutate(species = str_replace_all(Species, '_', ' ')) |> 
  left_join(metadata_tree_raw |> select(id, species)) |> 
  mutate(
    genome_mb = Length_Genome/1e6,
    perc_intronic = 100*Intronic/genome_mb,
    perc_exonic = 100*Exonic/genome_mb,
    perc_intergenic = 100*Intergenic/genome_mb,
  )

# abundance

total_metag_reads <-
  read_tsv("../../data/metag/samples_metag_reads.tsv", col_names = c("sample", "total_reads"))

sags_otutab <-
  read_tsv("../../data/metag/otutab_sags_100id_tax_summary.tsv") |>
  pivot_longer(names_to = "sample", values_to = "abundance", -1) |>
  left_join(total_metag_reads) |>
  mutate(perc = 100 * abundance / total_reads) |>
  select(-total_reads)

sags_mean_abundance <-
  sags_otutab |>
  filter(!str_detect(lca_taxonomy, "^lca:")) |>
  mutate(sag_id = paste0(str_remove(lca_taxonomy, ".*;"), "_", str_remove(lca_taxonomy, ";.*"))) |>
  group_by(sag_id) |>
  summarise(mean_abun = mean(perc)) |>
  left_join(sags_metadata |> select(sag_id, completeness, estimated_genome_size)) |>
  mutate(
    mean_abun_corrected = 100 * mean_abun / completeness,
    mean_abun_corrected_length = 1000*mean_abun_corrected/estimated_genome_size
    ) |>
  arrange(-mean_abun_corrected_length) |>
  mutate(
    Abundance =
      case_when(
        mean_abun_corrected_length >= 0.1 ~ "high",
        mean_abun_corrected_length >= 0.01 ~ "mid",
        TRUE ~ "low"
      ),
    Abundance = factor(Abundance, levels = c("high", "mid", "low"))
    )

sags_and_refs_genome_sizes <- 
  additional_genomes_busco |> 
  mutate(completeness = Complete + Fragmented) |> 
  left_join(additional_genomes_stats) |> 
  left_join(additional_genomes_architecture) |> 
  mutate(estimated_genome_size = 100*(sum_len/1e6)/completeness) |> 
  select(id, estimated_genome_size, completeness, starts_with('perc_')) |> 
  bind_rows(sags_est_genome_size) |> 
  select(-sag_id)

sags_clusters <- 
  read_tsv('../../data/sags/sags_clusters.tsv')

metadata_tree <- 
  metadata_tree_raw |> 
  left_join(sags_and_refs_genome_sizes) |> 
  left_join(sags_mean_abundance |> select(sag_id, Abundance)) |>
  left_join(sags_clusters, by = 'sag_id') |> 
  mutate(genome_available = !is.na(estimated_genome_size),
         cluster = factor(cluster))

pf_tree_rooted <-
  pf_tree |>
  root(pf_tree, node = 465)
```

## Functions

```{r}
base_tree_plotter <- function(tree, group_node, group_color, metadata_df = metadata_tree, group_name, treescale_x, treescale_y, legend_x, legend_y, add_cluster = F, cluster_offset = .012,show_abundance_legend = T, offset_labels = .02){
  
  # subset tree by node
  subset_tree <- 
    tree |> 
    extract.clade(group_node)
  
  # add metadata
  subset_tree %<>%
    dplyr::left_join(
      metadata_df |> mutate(label_tree = str_remove(name_long,str_c('_',group_name))),
      by = c("label" = "id")
    )
  
  p_base_tree <- 
    subset_tree |> 
    ggtree() + 
    geom_tiplab(
      parse = T,
      align = T,
      offset = offset_labels,
      aes(
          label = label_tree,
          color = genome_available
          ),
      linesize = 0.2
      ) +  
    scale_color_manual(values = c('gray70',group_color)) +
    guides(color = 'none', fill = guide_legend(override.aes = list(size = 4))) +
    geom_treescale(x = treescale_x, y = treescale_y) +
    ggnewscale::new_scale_color() +
    geom_tippoint(aes(color = Abundance, fill = Abundance), size = 3, shape = 21, na.rm = T) +
    scale_fill_manual(values = c("sienna", "khaki", "pink"), na.value = NA, na.translate = F) +
    scale_color_manual(values = rep('gray20',3), na.value = NA, na.translate = F) +
    theme(legend.position = 'inside',
          legend.position.inside = c(legend_x,legend_y),
          legend.background = element_blank()
          )
  
  if (show_abundance_legend == F){
    
    p_base_tree <- 
      p_base_tree +
      guides(fill ='none', color ='none')
    
  }
  
  if (add_cluster){
    
    cluster_colors <- c("#4dbbd5ff", "#e64b35", "#00a087", "#3c5488", "#f39b7f")
    
    p_base_tree <- 
      p_base_tree +
      ggnewscale::new_scale_fill() +
      ggnewscale::new_scale_color() +
      geom_tippoint(aes(color = cluster, fill = cluster, x = x+cluster_offset), size = 3, shape = 22, na.rm = T) +
      scale_color_manual(values = rep('gray20',5), na.value = NA, na.translate = F) +
      scale_fill_manual(values = cluster_colors, na.value = NA, na.translate = F) +
      guides(color = 'none', fill = guide_legend(override.aes = list(size = 4)))
    
  }
  
  return(p_base_tree)
}

genome_size_and_introns_plotter <- function(base_plot_tree, group_color, metadata_df = metadata_tree){
  
    order_labels <- 
      get_taxa_name(base_plot_tree)
  
    p_genome_sizes <- 
      metadata_df |> 
      filter(id %in% order_labels) |> 
      mutate(id = factor(id, levels = order_labels)) |> 
      ggplot(aes(x = estimated_genome_size, y = fct_rev(id))) +
      geom_col(color = 'gray20', linewidth = 0.2, fill = alpha(group_color,0.5)) +
      theme_sags() +
      scale_x_continuous(expand = c(0,0)) +
      labs(x = 'Est. genome size (Mbp)',  y = NULL) +
      theme(
        axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        panel.border = element_blank(),
        axis.line.x = element_line(linewidth = 0.4, color = 'gray20')
    )
    
    p_perc_intronic <-
      metadata_df |> 
      filter(id %in% order_labels) |> 
      mutate(id = factor(id, levels = order_labels)) |> 
      ggplot(aes(x = perc_intronic, y = fct_rev(id))) +
      geom_col(color = 'gray20', linewidth = 0.2, fill = alpha(group_color,0.2)) +
      theme_sags() +
      scale_x_continuous(expand = c(0,0)) +
      labs(x = 'Intronic regions (%)',  y = NULL) +
      theme(
        axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        panel.border = element_blank(),
        axis.line.x = element_line(linewidth = 0.4, color = 'gray20')
    )
    
    plot <- 
      wrap_plots(
        p_genome_sizes,
        p_perc_intronic
      )
      
    return(plot)
}

genome_architecture_plotter <- function(base_plot_tree, group_color, metadata_df = metadata_tree, add_clusters = T ,spacer_width = -0.09){
  
    order_labels <- 
      get_taxa_name(base_plot_tree)
  
    architectures_df <- 
      metadata_df |> 
      filter(id %in% order_labels) |> 
      select(id, estimated_genome_size, starts_with('perc_')) |> 
      pivot_longer(cols = -c(id,estimated_genome_size), names_prefix = 'perc_', names_to = 'type', values_to = 'perc') |> 
      mutate(length = perc*estimated_genome_size/100,
             id = factor(id, levels = order_labels),
             type = factor(str_to_sentence(type), levels = c('Exonic','Intronic','Intergenic')))
    
    p_architectures <- 
      architectures_df |> 
      ggplot(aes(x = length, y = fct_rev(id))) +
      geom_col(color = 'gray20', linewidth = 0.2, fill = group_color, aes(alpha = fct_rev(type))) +
      scale_alpha_manual(values = c(0.333, 0.6666, 1), name = NULL, 
                         guide = guide_legend(reverse = T, override.aes = list(fill = 'gray30'))) +
      theme_sags() +
      scale_x_continuous(expand = c(0,0)) +
      labs(x = 'Est. genome size (Mbp)',  y = NULL) +
      theme(
        axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        legend.text = element_text(size = 9),
        panel.border = element_blank(),
        legend.title = element_text(size = 10),
        axis.line.x = element_line(linewidth = 0.4, color = 'gray20'),
        
    )
    
    if (add_clusters){
      
      cluster_colors <- c("#4dbbd5ff", "#e64b35", "#00a087", "#3c5488", "#f39b7f")
      
      p_clusters <-
        metadata_df |> 
        filter(id %in% order_labels) |>
        mutate(id = factor(id, levels = order_labels)) |> 
        ggplot(aes(x = '', y = fct_rev(id))) +
        geom_tile(aes(fill = cluster), color = 'white', linewidth = 0.2) +
        theme_void() +
        theme(
          legend.text = element_text(size = 9),
          legend.title = element_text(size = 10)
        ) +
        guides(color = 'none', 
               fill = guide_legend(title = 'Cluster')) +
        scale_fill_manual(values = cluster_colors, na.value = NA, na.translate = F)
      
      p_architectures <- 
        wrap_plots(
          p_clusters,
          plot_spacer(),
          p_architectures,
          widths = c(.08,spacer_width,.92),
          guides = 'collect'
        ) &
        theme(legend.position = 'top', legend.box = 'vertical')
      
    }
    
    
    return(p_architectures)
  
}
```

## Specific clades

### Prymnesiophyceae

```{r}
group_name = 'Prymnesiophyceae'
tree = pf_tree
group_node = 611
group_color = 'seagreen'
treescale_x = 0
treescale_y = 32
legend_x = 0.09
legend_y = 1

p_prym_base_tree <- 
  base_tree_plotter(tree, group_node, group_color, metadata_df = metadata_tree, group_name, treescale_x, treescale_y, legend_x, legend_y)

p_prym_architectures <- 
  genome_architecture_plotter(p_prym_base_tree, group_color,add_clusters = T)

p_prym_tree_with_architectures <- 
  wrap_plots(
    p_prym_base_tree + xlim(0,.8),
    plot_spacer(),
    p_prym_architectures,
    widths = c(.8,-.24,.2)
  )
```

### Chlorarachnea

```{r}
group_name = 'Chlorarachnea'
tree = pf_tree_rooted
group_node = 893
group_color = 'deeppink3'
treescale_x = 0
treescale_y = 3.5
legend_x = 0.09
legend_y = .8

p_chlor_base_tree <- 
  base_tree_plotter(tree, group_node, group_color, metadata_df = metadata_tree, group_name, treescale_x, treescale_y, legend_x, legend_y, show_abundance_legend = F, offset_labels = 0.025)

p_chlor_architectures <- 
  genome_architecture_plotter(p_chlor_base_tree, group_color, add_clusters = T)

p_chlor_tree_with_architectures <- 
  wrap_plots(
    p_chlor_base_tree + xlim(0,1),
    plot_spacer(),
    p_chlor_architectures,
    widths = c(.8,-.23,.2)
  )
```

### Chrysophyceae

```{r}
group_name = 'Chrysophyceae'
tree = pf_tree
group_node = 750
group_color = "goldenrod3"
treescale_x = 0
treescale_y = 16
legend_x = 0.09
legend_y = .9

p_chryso_base_tree <- 
  base_tree_plotter(tree, group_node, group_color, metadata_df = metadata_tree, group_name, treescale_x, treescale_y, legend_x, legend_y, show_abundance_legend = F, offset_labels = 0.04)

p_chryso_architectures <- 
  genome_architecture_plotter(p_chryso_base_tree, group_color, add_clusters = T,spacer_width = -.085)

p_chryso_tree_with_architectures <- 
  wrap_plots(
    p_chryso_base_tree + xlim(0,1.5),
    plot_spacer(),
    p_chryso_architectures,
    # widths = c(.8,-.25,.2)
    widths = c(.8,-.24,.2)
  )  
```

### Merged plot

```{r}
fig <- 'S6'
caption <- get_caption(fig)

figS6 <- 
  wrap_plots(
    p_prym_tree_with_architectures,
    p_chryso_tree_with_architectures & guides(fill = 'none', color = 'none', alpha = 'none'), 
    p_chlor_tree_with_architectures & guides(fill = 'none', color = 'none', alpha = 'none'),
    heights = c(.66,.26,.075)
  ) +
  plot_annotation(tag_levels = list(c('A','','','B','','','C')))  

figS6_with_caption <-
  figS6 +
  plot_annotation(caption = caption) &
  theme(
    plot.caption = element_textbox_simple(
      margin = margin(5, 0, 5, 5),
      width = unit(1, "npc"),
      size = 10
    )
  )

ggsave(plot = figS6 , filename = '../../figures/supp_figures/figS6.pdf', height = 15, width = 10)
ggsave(plot = figS6_with_caption, filename = '../../figures/supp_figures/with_caption/figS6_with_caption.pdf', height = 16, width = 10)
```

